---
title: "Vectors and Matrix"
author: "Eralda Gjika"
date: "1/5/2022"
output:
  html_document:
    df_print: paged
---

Instructions: Ctrl+Alt+I how to Create a chunk

```{r}
3+5
log(8)+sin(pi)
```

```{r}

```

# Start

In this material you will have the opportunity to learn and test your self with: R basics, data structures, vectors, matrices and dataframe. As well as the fundamentals of R programming: Control structures and writing functions.

##### How to use Help? 

You can start by typing for example if you want info about "statistics" use:

??statistics 

or

help.search(stat)


Otherwise you can use the Help menu in R or Rstudio.
#### Operators

##### Arithmetic and Logic

###### Arithmetic

**addition +**

**subtraction -**

**multiplication * **

**division /**

 
#### Logic:

**< less than**

**> greater than**

**<= less than or equal**

**>= greater than or equal**

**== equivalent**

**!= different**

 
#### Example
```{r}
10-8
3+5
4+6*2
1-3*(4-8)/5
```
Assign values to a parameter:
```{r}
a<-7 
a
b=9
b
A<-a
A
# g
```

## NOTE! 
### There are some restrictions on object naming:

• May not contain "weird" symbols such as, +, -, #.
• A period (.) And a hyphen (_) are allowed, as well as names beginning with a period (.)
• A name cannot start with a number

```{r}
crystal.1<-6
crystal.1
mexico_dsde<-"SUPER"
mexico_dsde
# 2.crystal<-12 # comment 
# crystal-2
# crystal
#  ana crystal<-3
# crystal_3
```

#### scan()
The scan () function makes it possible to insert some data in R:

```{r}
#crystal<-scan()
# crystal
```

### VECTORS
A vector in R can be declared in several ways. 
You can use: 1) <- or 2) =, and immediately after it c ().
Try writing:
```{r}
a=c(3,4,5,6,3)
a
ba<-c(3,7,5,6,13)   
ba
t(a)
t(ba)
t(t(a))
a+3
a-1
a/5
a*8
```
Arithmetic calculations with vectors:
```{r}
a+ba
a-ba
a/ba
a*ba
length(a)
length(ba)
D<-c(3,4,6)
a+D
```
#### Minimum and Maximum
In R, we can find the minimum or maximum element of a vector using the commands min () or max ():
```{r}
min(a)
max(ba)
min(a,ba)
max(a,ba)
f<-c(min(a),min(ba))
f
# How can you do it for max()
```
Parallel minima and maxima: pmin() and pmax() finds the minimum (max) value in the same position
```{r}
w<-c(2,4,3,2,1)
u<-c(8,5,3,2,5)
s<-c(0,3,6,81,23)
pmin(w,u,s)
pmax(w,u,s)
# Try to add an element to vectors and re-call the functions, what you observe?
```
#### Sort and Order
sort () and order ()
The sort () function makes it possible to sort the elements of a numeric vector from the smallest value to the largest value:
```{r}
sort(a)
sort(ba)
```
The order () function makes it possible to sort the values of a vector but careful , displays their positions:
```{r}
a
order(a)
ba
order(ba)
```
#### Different length of vectors
What happens in R if the vectors have different lengths?
```{r}
d=c(1,2,3,4,2,3,1,0)
e<-c(3,4,5,6,3)
d+e
d-e
d/e
e*d
```
### DESCRIPTIVE STATISTICS
```{r}
length(a)
sum(a)# sum
mean(a)# arithmetic mean
mean(a,trim=1)# trim mean
var(a)# variance
sd(a)# standard deviation
range(a)# range
quantile(a)# quantiles
IQR<-c(quantile(a,0.25),quantile(a,0.75))# inter quartile range 
fivenum(a)# fivenum
library(moments)
skewness(a)
kurtosis(a)
```
EXERCISE: Write in R some commands that make it possible to calculate the dispersion of a vector with discrete data.
##### Missing data
```{r}
x=c(5,7,4,1,2,8,9,6,3,2,9,NA,4,5,3,NA, 4, 68, NA,32)
mean(x)
mean(x,na.rm=TRUE)
```
Can we change values to the elements of a vector? Of course yes!
```{r}
x
x[3]=0
x
x[c(1,5)]=0
x
x[4]=mean(x)
# elements in sequence from position 1 to 4 ?
# vector x by removing element in position 5
x2<-x[-5]
x2
# remove elements in position 3,7,8
# x[c(-,-,-)]
# a shorter vector from x
x3<-x[3:8]
x3
index<-c(3,6,7)
x[index]=NA
x
sum(a<5)
sum(a[a<5])
sum(a*(a<5))
#you have same results. Why? Let us go step by step.
a<5
1*(a<5)
a*(a<5)  
sum(a*(a<5) )
```
#### EXERCISE
Find the mean of all elements in vector a smaller than 2?
```{r}
mean(a<2)
```
#### EXERCISE
What do the following commands display in R editor? Describe in words the result.
```{r}
c=sum(a,a<5) 
c
d=sum(a,a<2)
d
sum(b>3|b<6)
sum(b>3 & b<6)
```
#### Operator = = 
This operator "==" tests all elements of the vector if they are equal to the declared value. R responds with a vector with TRUE or FALSE values in the controlled positions.
```{r}
x=c(5,7,4,1,2,8,9,6,3,2,9,NA,4,5,3,NA, 4, 68, NA, 32)
x==3
is.na(x)# gives TRUE or FALSE for each element of vector x
which(x==68)
# what about: which(x==max(x)) consider to remove the NA values
which(is.na(x))# gives the position of NAs in vector x
```
#### Match 
Match function - order is important.
Where do the values in the second vector appear in the first vector?’
```{r}
d=c(1,2,3,4,2,3,1,0)
e<-c(3,4,5,6,3)
match(d,e)
match(e,d)
# note that it shows the first position where the value is observed
```
#### Absolute value
The function of finding the absolute value of a number in R is abs ():
```{r}
yy=c(-1,2,-3,1,-4,5,3,55)
abs(yy)
# combining functions!
which(abs(yy-2)==min(abs(yy-2)))
all(yy>0) # are all elements in yy greater than 0?
any(yy>0) # is any element in yy greater than 0?
```
### SEQUENCE
If we have two given values (g <f or f <g) and we want to create a vector of values from g to f or from f to g, moving by 1 unit then we use the symbol:
The function seq () gives a range of values according to some conditions e.g.
```{r}
e1=19
e2=1
seq(e1,e2)
seq(e1,e2,-1)
# seq(e1,e2,2)# put a minus before 2
seq(2,13,3)
seq(13,2)#? correct it
i=1:15
i
j=i:(i+1)
j
```
#### EXERCISE
Construct a sequence of an arithmetic progression with d=4, 20 values , starting from 0.
#### TABLE -function
Helps organize the observation in a frequency table.
```{r}
table(x)
table(x)/length(x)
# what if we want it in %?
table(x)*100/length(x)
```
#### Categorical variable 
```{r}
var<-c("yes","no")
v2<-c("y","n","n","n","y","n") # table it
table(v2)/length(v2)
crystal<-c("ana","eralda","daniel","data science")
crystal
length(crystal)
nchar(crystal)
sum(nchar(crystal))
paste(var,crystal,sep=" ")
```
#### Extract in strings
Extracting parts of strings
```{r}
phrase<-"the student of data science course are fantastic"
q<-character(sum(nchar(phrase)))
for (i in 1:sum(nchar(phrase))) q[i]<- substr(phrase,1,i)
q
strsplit(phrase,split=character(0))
```
#### Factors
Qualitative data are often used to classify data at different levels or factors (modalities). To create a factor (modality) in R is simple with the command factor or as.factor. See the following example:
```{r}
v2
factor(v2)
table(factor(v2))
is.numeric(x) 
v3<-"crystal"
class(v3) 
is.character(a) 
v4<-"system"
paste(v3,v4)
paste(v3,v4,sep="")
substr(v3,2,5)
```
#### Sequence 
We can use the function seq () to count deciles:
```{r}
decile<-seq(0,1,0.1)
Decile=quantile(yy,decile)
Decile
```
#### Repeat
• The rep () function repeats a parameter, value, value vector several times, for example:
```{r}
rep("A",10)
rep(x,each=3)
rep(2:5,c(4,1,4,2))# repeat each element with the given frequency
```
#### Sort
• The sort () function sorts the values of the numeric vector, for example:
```{r}
sort(x)
(sort(x))^(-1) 
# let's combine functions!
rev(sort(x))
rev(sort(x))[1:3]
sum(rev(sort(x))[1:3])
```
#### Round
The function that allows a rounding to R is: round ().
```{r}
round(1.34)
round(1.345,2)
```
#### Sampling in R
```{r}
sample(1:50,6) 
sample(1:50,6)
sample(1:50,26,replace=TRUE)
sample(c("H","T"),15,replace=TRUE)# coin toss
sample(c("H","T"),15,replace=TRUE,prob=c(0.7,0.3))
```

#### EXERCISE 1
Simulate the roll of a dice 20 times.

#### EXERCISE 2
For the following data, calculate: arithmetic mean, mean, maximum value, minimum value, amplitude, mean square deviation, quantiles.
45, 43, 46, 48, 51, 46, 50, 47, 46, 45,54,76,13,89,8,

#### EXERCISE 3
The following values are given:
653 656 659 662 664 668 671 674, 564, 783, 943, 350, 567, 876
Write these numbers in R. Use the diff () function in the data. What do you get?
Try: diff(diff())?

#### EXERCISE 4
Suppose you work for 10 days and your working hours are below:
   17 16 20 24 22 15 21 15 17 22
a) Construct a vector with the data in R.
b) Find the maximum number of working hours.
c) Which day has been the most tiring, so that has lasted the longest for you?
d) On average, how many working hours have you had these 10 days? On which day did you work the least?
e) Value 24 was incorrect. It should have been 18. How would you fix this? Once you have adjusted find the new average.
f) How many times was your working time 20 hours or more? To answer this question try: sum (vector> = 20)
(if you called the working hours vector "vector") What do you get? What percentage of the days did you work less than 17 hours? 


#### MATRIX
Union and intersect of two sets A and B
```{r}
A=c(1,4,2,1,2,7)
B=c(1,3,5,6,1,2,8,4)
union(A,B)
intersect(A,B)
```

#### setdiff() and setequal()
```{r}
setdiff(A,B) # elements of A which are not in B
setdiff(B,A) #elements of B which are not in A
setequal(A,B) # TRUE if the sets are equal 
```
#### %in% 
controles if a set X is in Y, returns TRUE or FALSE for elements in Y which are or not in X (pay attention to the order)
```{r}
X<-1:12
Y<-5:9
Y %in% X
X %in% Y
all(X%in%Y)# if we want to test all elements of X if they are in Y
```
#### Names of rows and columns in a matrix
```{r}
M<-matrix(c(1,0,4,0,1,0,7,9,1),nrow=3)
M
rownames(M)<-c("A","B","C")
M
colnames(M)<-colnames(M,do.NULL=FALSE,prefix="Var.")
M
```
Or, we can use dimnames argument :
```{r}
D<-matrix(c(1,2,3,11,12,13), nrow=2, ncol=3, byrow=TRUE, dimnames=list(c("row 1", "row 2"), c("Col 1", "Col 2", "Col 3")))
D
```

##### Abbreviations
Also we can name the rows and columns by letters or month name or abbreviations for months:
```{r}
rownames(D)=LETTERS[1:2]
D
colnames(D)=letters[1:3]
D
colnames(D)=month.name[1:3]
D
colnames(D)=month.abb[1:3]
D
```

#### Matrix Dimenssions
We can obtain the number of rows and columns using:
```{r}
nrow(D)
ncol(D)
dim(D)
```
#### Grouping in Matrix
##### Exercise
Let suppose each row is an individual (item, observation)
```{r}
M1<-matrix(c(1,0,4,0,1,0,7,9,1,4,6,2),nrow=4)
group=c("G1","G2","G2","G1")
rowsum(M1, group)
M1
tapply(M1, list(group[row(M1)], col(M1)), sum)
M1
aggregate(M1,list(group),sum)
M1
cols<-apply(M1,2,mean)# apply to the columns the mean
M1
```
##### cbind and rbind
Add rows or columns using cbind and rbind
```{r}
M2<-matrix(c(1,0,4,0,1,0,7,9,1,4,6,2,5,0,7,8),nrow=4)
M2
M2<-rbind(M2,apply(M2,2,mean))# apply to the column the mean
M2<-cbind(M2,apply(M2,1,var))# apply to the row the variance
colnames(M2)<-c(1:4,"variance")
rownames(M2)<-c(1:4,"mean")
M2
```
#### Matrix multiplications
```{r}
M2<-matrix(c(1,0,4,0,1,0,7,9,1,4,6,2,5,0,7,8),nrow=4)
M2%*%M2	# M2^2 quadratic matrix of M2
```
#### Solving linear equations
```{r}
 solve(M2)# gives the inverse of matrix M2
```
#### Example
let supose we have the equations: 4x+5y=1 and 6x+7y=9 and we want to find the solutions x and y. 
```{r}
A<-matrix(c(4,5,6,7),nrow=2,byrow=TRUE)
A
b<-c(1,9)
b
solve(A,b)# the solution is x=19 and y=-15
```
#### Other helpful functions for matrices are:
t(), det(), qr(), chol(), eigen(), prcomp(), diag(), array(), outer()
```{r}
M2<-matrix(c(1,0,4,0,1,0,7,9,1,4,6,2,5,0,7,8),nrow=4)
M2
diag(M2)# the diagonal of M2
det(M2)# the determinant of matrix M2
t(M2)# the transpose of M2
```
### More computations with matrices
```{r}
library(dplyr)
library( miscTools)
```

```{r}
x <- matrix(rnorm(20), nrow = 5, ncol = 4)
print(x)
```
### Row averages
```{r}
print(rowMeans(x))
print(rowMedians(x))
```
### Column averages
```{r}
print(colMeans(x))
print(colMedians(x))
```
### Row variabilities
```{r}
library(matrixStats)
print(rowVars(x))
print(rowSds(x))
print(rowMads(x))
print(rowIQRs(x))
```
### Column variabilities
```{r}
print(rowVars(x))
print(colSds(x))
print(colMads(x))
print(colIQRs(x))
```
### Row ranges
```{r}
print(rowRanges(x))
print(cbind(rowMins(x), rowMaxs(x)))
print(cbind(rowOrderStats(x, which = 1), rowOrderStats(x, which = ncol(x))))
```
### Column ranges
```{r}
print(colRanges(x))
print(cbind(colMins(x), colMaxs(x)))
print(cbind(colOrderStats(x, which = 1), colOrderStats(x, which = nrow(x))))
```

```{r}
x <- matrix(rnorm(2400), nrow = 50, ncol = 40)
```
### Row standard deviations
```{r}
d <- rowDiffs(x)
s1 <- rowSds(d) / sqrt(2)
s2 <- rowSds(x)
print(summary(s1 - s2))
```
### Column standard deviations
```{r}
d <- colDiffs(x)
s1 <- colSds(d) / sqrt(2)
s2 <- colSds(x)
print(summary(s1 - s2))
```

## DATAFRAME
How to create a data frame? 
We can use the function data.frame(). If we have two variables let's suppose age and high of some individuals then we can create a dataframe with this two vectors using the function as below. Pay attention to the fact that the vectors should have the same length. NA-s are considered missing values but are counted as missing observations. We can use function na.omit() or the argument na.rm=TRUE inside a function to deal with missing values. 
Below are examples of some functions used in dataframe-s: 

-> subset() will create a subset of a given dataframe
-> transform()  will add a variable to the dataframe
-> split()  will split the variables from the dataframe

```{r}
age=c(60,62,57,72,66,20,45, 89, 67,45 )
height=c(170,166,158,180,165,145,165,157,190,187)
i<-seq(1,10)
df=data.frame(age,height)
df
df$age
df$height
df$i
df$age[height<=170]
height<=170 # attention to the brackets
df$height[df$age<=65 & df$age>60]
df.1=subset(df,df$age<70)# create a subset of df of those individuals age <70
df.2=transform(df,log.age=log(df$age))# add a column with log(age)
df.2
# edit(df.2)# edit the dataframe
df.s=split(df$age,df$height)
df.s
lapply(df,mean)
sapply(df,mean)
lapply(df,var)
sapply(df,median)
lapply(df,log)
sapply(df,exp)
str(df)# structure of the dataframe
name=c("Ana","Ben","Silva","Etta","Tim","Noah","Emia","Eino","Emanuel","Christina")
names(age)<-name
age
```


## Appendix
### ARRAYS
```{r}
A<-letters[1:24]# 24 letters of the alphabet
dim(A)=c(3,2,4)# create an array with matrices 3x2, and 4 matrices (6x4=24)
A
A[,,1:3]# select all the letters from a to r
```
##### LISTS
```{r}
crystal_list=list(Name="Ana" , Salary=1000 , City="Bucurest")
typeof(crystal_list)
length(crystal_list)
names(crystal_list)
unlist(crystal_list)
crystal_list[3]
crystal_list[[1]]# information about 1st variable
crystal_list$Name
crystal_list[1:2]
crystal_list$Age=25
crystal_list
crystal_list=list(Name=c("Ana","Ben"), Salary=c(1200,5600), City=c("Bucurest", "Tirana"))
summary(crystal_list)
summary(crystal_list$Salary)
crystal_list.1=list(Gender=c("F","M"),Profession=c("student","Data Scientist"))  
crystal_final=list(crystal_list,crystal_list.1)# create a new list with additional information from another list of same individuals
```



End!

Now is your turn to try your exercise!